---
title: "STAT 3340 Assignment 4, Fall 2018"
author: "James MacPhee"
date: 'Banner:  B00768516'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

1.
```{r}
yield=c(123,128,166,151,156,150,178,125,112,174,187,117,100,116,153,155,
168,109,195,158,135,175,140,167,130,132,145,183,176,120,159,142,120,187,
131,167,155,184,126,168,156,186,185,175,180,138,206,173,147,178,188,154,
146,176,165,191,193,190,188,169)
crop=as.factor(rep(c("W","C","S","R"),15))
fertilizer=as.factor(c(rep("X",20),rep("FS",20),rep("SS",20)))
```

  a)
  
```{r}
boxplot(yield~crop)
boxplot(yield~fertilizer)
```
The relationships between yield and fertilizer seem to be more correlated and normally distributed than yield and crop. Crops C and R are quite skewwed while all three of the fertilizers seem to be relatively unskewwed.
  b)
  
```{r}
IW=ifelse(crop=="W",1,0)
IC=ifelse(crop=="C",1,0)
IS=ifelse(crop=="S",1,0)
IX=ifelse(fertilizer=="X",1,0)
IFS=ifelse(fertilizer=="FS",1,0)
```
  
  c) 
  $y = \beta_0 + \beta_1IW + \beta_2IC + \beta_3IS + \beta_4IX + \beta_5IFS + \beta_6IWIX + \beta_7IWIFS + \beta_8ICIX + \beta_9ICIFS + \beta_{10}ISIX + \beta_{11}ISIFS + \epsilon$

  
  d) 
  
```{r}
ls=lm(yield~IW+IC+IS+IX+IFS+IW*IFS+IC*IFS+IS*IFS+IW*IX+IC*IX+IS*IX)
anova(ls)
```
 
 
  e)
    i)  
  $H_0:\beta_6=\beta_7=\beta_8=\beta_9=\beta_{10}=\beta_{11}=0$
  $H_a:$ At least one of $\beta_6,\beta_7,\beta_8,\beta_9,\beta_{10},\beta_{11}\neq0$
  
  ii)
  $y=\beta_0+\beta_1IW + \beta_2IC + \beta_3IS + \beta_4IX + \beta_5IFS + \epsilon$
  
  iii)
```{r}
lsred=lm(yield~IW+IC+IS+IX+IFS)
anova(lsred)
```  
  iv) $ $
```{r}
F=((27446.3-21220.4)/6)/442.1
F
```
  v) $ $
```{r}
p=1-pf(F,6,48)
p
```
  Using the table from the class website the p-value is between .01 and .05
 
 
 
  2.
   a)
    i)
    $F_{obs}=(((62.12+0.350+0.796)-62.120)/2)/0.441=1.2993197$
    ii)
    $F_{obs}$~$F_{2,143}$
```{r}
F_obs=((63.266-62.12)/2)/0.441
p_value=1-pf(F_obs,2,143)
p_value
```
  
  b)
  95% CI for cement: $(0.010236-(1.96)(0.001433),0.010236+(1.96)(0.001433))=(0.00742732,0.01304468)$
    
    
3. 
```{r}
fish=read.csv('http://bsmith.mathstat.dal.ca/stat3340/Data/fish.csv',header=T)
age=fish[,1]
temp=fish[,2]
length=fish[,3]
```
  a)
```{r}
lm.fish=lm(length~age+temp,data=fish)
summary(lm.fish)
```

  b)
```{r}
lm.fishQ=lm(length~age+temp+age*temp+I(age^2)+I(temp^2),data=fish)
summary(lm.fishQ)
```
  c)
```{r}
resid.fishQ=resid(lm.fishQ)
fitted.fishQ=fitted(lm.fish)
plot(fish,resid.fishQ)
qqnorm(resid.fishQ)
```
  To me the quadratic model seems appropriate. There isn't really any tailing going on or any major skewness. Nothing outside of randomness.
  
  d)
  i)
  $H_0: \beta_3=\beta_4=\beta_5=0$
  $H_a: \beta_3,\beta_4,$or$\beta_5\neq0$
  
  ii)
```{r}
anova(lm.fish)
anova(lm.fishQ)
Fobs=((14759878-2150908)/3)/56603
Fobs
```
  iii)
```{r}
p.value=1-pf(Fobs,3,38)
p.value
```
  e)
    i)
```{r}
summary(lm.fish)
anova(lm.fish)
e1=resid(lm.fish)
```
    ii)
```{r}
lm.age=lm(I(age^2)~age+temp)
e2=resid(lm.age)
```
    iii)
```{r}
lm.resids=lm(e1~e2-1)
anova(lm.resids)
```
    iv)
```{r}
lm.full=lm(length~age+temp+I(age^2))
summary(lm.full)
anova(lm.full)
```
    v)
```{r}
lm.full$coefficients
lm.resids$coefficients
```
    vi)
```{r}
coef(lm.fish)+coef(lm.resids)*(-coef(lm.age))
lm.full$coefficients
```
    vii)
```{r}
anova(lm.full)$Sum
anova(lm.resids)$Sum
```
    As you can see the numbers for SSE are the same in both models: 6372687
    
    viii)
```{r}
anova(lm.full)$Sum
anova(lm.resids)$Sum
```
    Just like above you can see that the values are the same, except this time instead of looking at SSE we are interested in the SSR (age^2 portion at least): 8387191
    
    ix)
```{r}
plot(e2,e1,main="Added variable plot for age^2")
abline(lm.resids)
```
  It doesn't seem appropriate to add in the quadratic term now because the data doesn't really follow any trend at all.
  
  
4.
  a) Answer: yes
  b) Answer: no
  c) Answer: yes
  d) Answer: no


5.
  a)
  You would compare models 3 and 6.
  $Fobs=((180-120)/6)/1.31868=7.583341$~$F_{6,91}$

  b)
  You would compare models 4 and 6.
  $Fobs=((140-120)/3)/1.237113=5.388891$~$F_{3,91}$
  
  c)
  Obviously compare models 4 and 5. You can't use the F-test because it would lead to a negative value. It is quite hard to compare when you are adding and removing terms  between the two models.


6.
$y=weight$ , $x_1=age$ , $x_2=1 if male and x_2=0 if female$
$y=\beta_0+\beta_1x_1+\beta_2x_1^2+\beta_3x + \beta_4x_2^2+\beta_5x_1x_2+\epsilon$
I think this is all the question is asking. I am confused by the wording.
