---
title: "STAT 3340 Assignment 1, Fall 2018"
author: "James MacPhee"
date: "B00768516"
Banner:  "09/24/2018"
output:
  word_document: default
  pdf_document: default
---
======================================================================


1. Find the equation of the line which passes through the points (1,1) and (4,5).

   $$y=\frac{-1}{3} + \frac{4}{3} x$$

2. Suppose you are given three data points (1,4), (2,6) and (3,7) and the line $y = 1 + 4x$. 
Give the three residuals and their sum of squares.

    r_1 = -1, r_2 = -3, r_3 = -6
    $$\sum_{i=1}^3 r_i^2 = 46$$

3. Some data gives the summaries: $n = 10$, $\sum_{i=1}^{10}  x_iy_i = 100$, $\sum_{i=1}^{10} x_i = 20$ and $\sum_{i=1}^{10} y_i = 10$. 
   Suppose that the response $y$ is temperature in degrees Celcius. 

    + 3.a) $S_{xy}$= 80
    + 3.b) If the response was converted to temperature in degrees Fahrenheit $y'$, so that $y' = 32+1.8y$, what is $\sum_{i=1}^{10} y'_i$? 32+1.8(10) = 338
    + 3.c) If the response was converted to temperature in degrees Fahrenheit $y'$, so that $y' = 32+1.8y$, what is $S_{xy'}$? (32(20)+1.8(100))-(20)(338)/10 = 144


4. A random sample of 11 elementary school students is selected,
  and each student is measured on a creativity score (x) using a
  well-defined testing instrument and on a task score (y) using a new
  instrument. The task score is the mean time taken to perform several
  hand-eye coordination tasks. The data are as follows.
  
```{r}
 x=c(35,37,50,69,84,40,29,42,51,45,31)
 y=c(3.9,3.9,6.1,4.3,8.8,2.1,5.7,3.0,7.1,7.3,3.3)
 plot(x,y,xlab="Creativity score",ylab="Task score")
```

 Use R to do the following questions. Make sure to include your commands in the R markdown document.
 
   + 4a) Calculate the summaries $S_{xx}$, $S_{xy}$, $S_{yy}$ and $\bar{X}$ and $\bar{Y}$.
  
 
```{r}
N=length(x)
xbar=sum(x)/N ;  cat("xbar = ", xbar)
ybar=sum(y)/N ;  cat("ybar = ", ybar)
Sxx=sum((x-xbar)^2) ; cat("Sxx = ", Sxx)
Sxy=sum((x-xbar)*(y-ybar)) ; cat("Sxy = ", Sxy)  
Syy=sum((y-ybar)^2) ; cat("Syy = ", Syy)
```

   + 4b) Use these summaries to  calculate the least squares estimates of the  intercept and slope.
  
```{r}
 beta1 = Sxy/Sxx ; cat("beta_1 = ", beta1)
 beta0 = (ybar-beta1*xbar); cat("beta_0 =", beta0)
```
  
   + 4c) Calculate the predicted (fitted) values and the residuals.
   
```{r}
fit=beta0+beta1*x ; cat("Fitted Values: ", fit)
resids=y-fit ; cat("residuals: ", resids)
```
   +  4d) Plot the residuals (y axis) vs fitted values (x axis).


```{r}
 plot(fit,resids,xlab="Fitted Values",ylab="Residuals")
```

   + 4e) Do the residuals look random? If not, what does the pattern of the residuals suggest?
   
   + 4f) Calculate the mean of the residuals
to verify it is zero (to machine precision), and the correlation of the residuals with $X$ to verify it is also zero.  The R command to calculate
the correlation between $u$ and $v$ is cor(u,v).

```{r}
# 4e) Yes very random indeed
mean(resids)
cor(resids,x)
```

  + 4g) Calculate and print the error, regression and total sums of squares, using the data summaries.

```{r}
SST=Syy ; cat("SST = ", SST)
SSE= sum(resids)^2 ; cat("SSE = ", SSE)
SSR=SST-SSE ; cat("SSR = ", SSR)
```


5. The length of a species of fish is to be represented as a function
of the age and water temperature.  The fish are kept in tanks
at 25, 27, 29 and 31 degrees Celsius.  After birth, a test specimen
is chosen at random every 14 days and its length measured.  The data can be entered into R as

```{r}
fish=read.csv("http://bsmith.mathstat.dal.ca/stat3340/Data/fish.csv",header=T)
pairs(fish)
```
 
The pairwise plots don't really give an adequate picture of what is going on.
The following plot of length vs age, with different temperatures identified
suggests that growth is impeded by high temperatures.

```{r}

  color=rep("black",dim(fish)[1])
  color=ifelse(fish$temp==27,"blue", color)
  color=ifelse(fish$temp==29,"green",color)
  color=ifelse(fish$temp==31,"brown",color)
  legend.txt=c("25 degrees","27 degrees","29 degrees","31 degrees")

plot(fish$age,fish$length,col=color,ylab="length",xlab="age",cex=2,pch="*")
legend(100,2000,legend=legend.txt,text.col=c("black","blue","green","brown"))

```

The multiple regression model $length = \beta_0+\beta_1 age +\beta_2 temp +\epsilon$
was fit using R, as follows.


```{r}
fish.out=lm(length~age+temp, data=fish)
summary(fish.out)
anova(fish.out)
```

  +  Test for the overall significance of the regression.  
    
The hypotheses are: $H_0: \beta_1 = \beta_2 = 0$ vs
$H_A:\mbox{ one or more of } \beta_1, \beta_2 \ne 0$.

  + 5a) The observed value of the $F$ statistic is 84.98
  + 5b) The numerator degrees of freedom is 2
  + 5c) The denominator degrees of freedom is 41
  + 5d) Calculate the  p-value
```{r}
 1-pf(84.98,2,41)
```



  + 5e) Use the lm command to fit a model which includes only the variable age.

```{r}
ageonly.out = lm (length~age, data=fish)
summary(ageonly.out)
```

Test the appropriate hypothesis concerning "temperature".

  + 5f) state the hypotheses:  $H_0: \beta_1 = 0$  vs  $H_A: \beta_1 \ne 0$
    
  + 5g) Use the ANOVA command to compare the two models (here fish.out and ageonly.out)

```{r}
anova(fish.out, ageonly.out)
```

   + 5g(i)The observed value of the test statistic is 6.9202
   + 5g(ii) The numerator degrees of freedom is 41
   + 5g(iii)The denominator degrees of freedom is 42
   + 5g(iv) Calculate the p-value
    
```{r}
1-pf(6.9202,41,42)
```
    
    
  + 5h) The model 
     $$length = \beta_0+\beta_1 age +\beta_2 temp + \beta_3 age*temp +\epsilon$$
    includes the interaction between temperature and age.  The model
    is fit in R using "lm(length\~age+temp+age:temp)".  An equivalent
    model statement is "lm(length\~age*temp)".  Fit this model in R,
    and then use the "anova" command to carry out the F test of significance of the interaction.
    
```{r}
lm.interac=lm(length~age+temp+age:temp, data=fish)
anova(lm.interac,fish.out)
```
  + 5h(i) The null hypothesis is $H_0: \beta_3$ = 0  
  + 5h(ii) The alternative hypothesis is $H_A: \beta_3$ \ne 0
  + 5h(iii) The observed value of F is  3.8858
  + 5h(iv) The numerator degrees of freedom is 40
  + 5h(v) The denominator degrees of freedom is 41
  + 5h(vi) Calculate the p-value.

```{r}
1-pf(3.8858,40,41)
```
