---
title: "STAT 3340 Assignment 2, Fall 2018"
author: "James MacPhee"
date: 'Banner:  B00768516'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

For question 1, include the appropriate R code and process in markdown.

For questions 2  onward, you can either process in markdown, or include handwritten answers after printing.

A convenient processing option is to knit to html, then print to pdf from the browser's menu.
If you knit to Word, you will need to have appropriate fonts, etc, available in Word, in order to 
have the document rendered appropriately.  If you knit to pdf, line spacing may be an issue for questions 2-5.



1. The length of a species of fish is to be represented as a function
of the age and water temperature.  The fish are kept in tanks
at 25, 27, 29 and 31 degrees Celsius.  After birth, a test specimen
is chosen at random every 14 days and its length measured.


```{r}
data=read.csv("http://bsmith.mathstat.dal.ca/stat3340/Data/fish.csv",header=T)
age=data[,1]
temp=data[,2]
length=data[,3]
N=dim(data)[1]
y=length
X=cbind(rep(1,N),age,temp)
X[1:3,]
```

Fit the model $y = \beta_0+\beta_1 x_1 +\beta_2 x_2+\epsilon$ using matrix commands in R.

   + 1a) Calculate and display the 3x3 matrix $X'X$.

```{r}
xpx= t(X)%*%X
xpx
```
   + 1b) Calculate and display $[X'X]^{-1}$.

```{r}
xpxi= solve(xpx)
round(xpxi,5)
```

   + 1c) Calculate and display $\hat \beta$.
   (you can check your results with "lm(length~age+temp)"   

```{r}
betahat = xpxi%*%t(X)%*%y

round(betahat,3)
lm(length~age+temp)
```


   + 1d) Calculate the predicted values, the residuals, and the error sum of squares.
Print the error sum of squares.

```{r}
yhat = X%*%betahat
resids = y-yhat
SSE = t(y)%*%y - t(betahat)%*%t(X)%*%y
SSE
```
+ 1e) Use the matrix formula $SSR=\hat beta ' %*% X' %*% y - N {\bar y}^2$ to calculate the
regression sum of squares, and verify that SST=SSE+SSR, up to rounding error. 


```{r}
SST=sum((y-mean(y))^2)
SSR= t(betahat)%*%t(X)%*%y-N*mean(y)^2
SSE=SST-SSR
round(SST-SSE-SSR,3)
```
   
   +  Calculate, but do not print, the projection matrices $\mathbf{H}$, $\mathbf{I}-\mathbf{H}$, $\mathbf{J}/n$, and $\mathbf{I}-\mathbf{J}/n$

```{r}
k=2
H=X%*%xpxi%*%t(X)
I=diag(N)
ImH=I-H
ones=matrix(rep(1,N),byrow=T,ncol=1)
J=(ones%*%t(ones))/N
```

   + 1f)  Verify, by calculating their inner product, that $\mathbf{H}$ $\mathbf{y}$ and $(\mathbf{I}- \mathbf{H})\mathbf{y}$ are perpendicular. 
```{r}
t(H%*%y)%*%(ImH%*%y)
```

   + 1g)  Verify that $(\mathbf{J}/n)\mathbf{y} = \bar{y}\mathbf{1}$.

```{r}
diff= mean(y)*ones - J%*%y
round(diff, 5)
```


2.  For the data $y=(1,2,4)$ and $x=(0,1,2)$,

   + 2a)  To fit the model $y=\beta_0+\beta_1 x + \epsilon$
   using matrix calculations, what is the appropriate matrix $\mathbf{X}$?
   
<br>
1 0
<br>
1 1
<br>
1 2
<br>  
<br>  

   + 2b)  Calculate the two by two matrix matrix 
   $\mathbf{X'} \mathbf{X}$.
   
<br>  
3 3
<br>
3 5
<br>  
<br>  

   


   + 2c)  Calculate the  matrix 
   $[\mathbf{X'} \mathbf{X}]^{-1}$.
   
<br>  
5/6 -1/2
<br>
-1/2 1/2
<br>  
<br>  

   
 

   + 2d)  Calculate the  least squares estimates.
   
<br>  
betahat = (X'X)^(-1)X'y
<br>
X'y = [7 10]'
<br>
betahat = [5/6 3/2]'
<br>  
<br>  


3. Suppose $X$, $Y$ and $Z$ are random variables with means
$\mu_X=1$, $\mu_Y=2$ and $\mu_Z=3$, variances $\sigma^2_X = 9$,
$\sigma^2_Y = 4$ and $\sigma^2_Z = 1$, and covariances
$Cov(X,Y) = 3$, $Cov(Y,Z) = 1$ and $Cov(X,Z) = 1.5$. Let $U = X-Y+2Z$ and $V=X+2Y-2Z$.

   + 3a) What is $E(U)$?

   
<br>  
E[U] = E[X]-E[Y]+2E[Z] = 1-2+2(3) = 5
<br>  
<br>  

   + 3b) What is $Var(U)$?

   
<br>  
Var[U]=Var[X]+Var[Y]+4Var[Z]-2Cov[X,Y]+4Cov[X,Z]-4Cov[Y,Z]
<br>
Var[U] = 9+4+4(1)-2(3)+4(1.5)-4(1) = 13
<br>  
<br>  

   + 3c) What is $Cov(U,V)$?

   
<br>  
Cov[U,V] = Cov[X-Y+2Z,X+2Y-2Z] = Cov[X,X]+Cov[X,2Y]+Cov[X,-2Z]+Cov[-Y,X]+Cov[-Y,2Y]+Cov[-Y,-2Z]+Cov[2Z,X]+Cov[2Z,2Y]+Cov[2Z,-2Z]
<br>
Cov[U,V] = Var[X]+2Cov[X,Y]-2Cov[X,Z]-Cov[Y,X]-2Var[Y]+2Cov[Y,Z]+2Cov[Z,X]+4Cov[Z,Y]-4Var[X]
<br>
Cov[U,V] = 9+2(3)-2(1.5)-3-2(4)+2(1)+2(1.5)+4(1)-4(9) = -26
<br>  
<br>  



4.    In a study of 1006 men in four occupations, a multiple regression was carried out
to show how lung function was related to age, smoking and occupation.  The four
occupations represented in the study were physician, firefighter, farm worker,
and chemical worker.  The variables in the regression were:

$Y$ -  air capacity (ml) that the worker can expire in one second

$X_1$ - age in years

$X_2$ - number of cigarettes smoked per day

$X_3$ - 1 if chemical worker, 0 otherwise

$X_4$ - 1 if farm worker, 0 otherwise

$X_5$ - 1 if firefighter, 0 otherwise 


The model $Y=\beta_0+\beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_4 + \beta_5 X_5 + e$
was fit to the data with the following result:

$$\hat Y = 4500 - 39 X_1 -9 X_2 - 350 X_3 -380 X_4 -180 X_5$$

After fitting a linear regression model, it was found that $MSE=9$ and 
$$ ({\mathbf{X'X}})^{-1} = \left [ \begin{array}{c c c c c c}
    9 &   8  &  7  &  6  &  7   & 7 \\
    8 &   3  &  5  &  6  &  6   & 1 \\
    1 &   1  &  7  &  4  &  5   & 6 \\
    1 &   6  &  9  &  9  &  8   & 3 \\
    7 &   5  &  5  &  5  &  8   & 8 \\
    7 &   3  &  4  &  1  &  5   & 4 \\
\end{array} \right ]$$

   + 4a) Construct a 95\% confidence interval for $\beta_4$. 

   
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


   + 4b) Construct a 95\% prediction interval for $Y$, for a 30 year old 
physician who does not smoke.

   
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


   + 4c)  Write down a regression model which includes terms for the interaction
    of age and occupation, and the interaction of age and number of of
    cigarettes smoked per day.

   
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


5.  Suppose that $Y$ is a random vector with mean vector $(0,0,1)'$, and covariance matrix

$$  \left [ \begin{array}{c c c}
      1 \,\, 0 \,\, 0\\
      0 \,\, 2 \,\, 0 \\
      0 \,\, 0 \,\, 3   \end{array} \right ] $$

  Let $\mathbf b = (0,0,1)'$ and 

$$ \mathbf A = \left [ \begin{array}{c c c}
      1  \,\, 0  \,\, 0\\
      0  \,\,0   \,\,2 \\
      0  \,\, 1  \,\,0   \end{array} \right ] $$

   + 5a) What is the mean of $\mathbf A  \mathbf Y +  \mathbf b$?



   
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


   + 5b)    What is the covariance matrix of $\mathbf A \mathbf Y + \mathbf b$?


   
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


   + 5c)  What is the mean of $\mathbf Y' \mathbf A  \mathbf Y +  \mathbf b$?
    

   
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
